---
title: "Maddie's data analysis"
author: "Javier Tabima and Manning DelCogliano"
date: "2023-12-16"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading libraries

```{R}
library(dada2)
library(tidyverse)
library(googlesheets4)
library(viridis)
library(ggpubr)
```

# Creating in PS object

```{r}
gs4_deauth()
ww.metadata <- read_sheet(
  "https://docs.google.com/spreadsheets/d/1-R5HzaB2c5bHIbekrREiTmvWsC1Y-jhv5wHbcgmy5uc/edit#gid=51094388", sheet = 6
)
 
seqtab.nochim <- readRDS("/Smaug_SSD/MBB101/SaWheeler/nochimera.RDS")
taxa <- readRDS("/Smaug_SSD/MBB101/SaWheeler/taxa.RDS")

samples.out <- rownames(seqtab.nochim)
location <- gsub(samples.out, pattern = "\\d+.+", replacement = "", perl = T) # set location to be just the letters in the sample name
samdf <- data.frame(Location=location)
rownames(samdf) <- samples.out 

ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               sample_data(samdf), 
               tax_table(taxa))

dna <- Biostrings::DNAStringSet(taxa_names(ps)) #is this the sequences 
names(dna) <- taxa_names(ps) #is this the names of the sequences 
ps <- merge_phyloseq(ps, dna) 
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps))) #assign ASV names to DNA sequences within each sample to make things succinct
ps


to_remove <- grep(x=sample_names(ps), pattern = "^M\\d+|^SBFM\\d+|^UP\\d+|^W\\d+|^WN4.raw|PR21B.raw|PR21B|PR22B|CR11B|BB24B|CP24B|CP32B|BB14A|MB12B|MB14B|BB23A|CP42ASed.raw|PR51J.raw", perl = T, value = T) 

ps_ww <- prune_samples(sample_names(ps)[!(sample_names(ps) %in% to_remove)], ps)


ps_ww@sam_data$Site <- ww.metadata$Site[match(rownames(ps_ww@sam_data),ww.metadata$SampleName)]

ps_ww@sam_data$SampleType <-ww.metadata$SampleType[match(rownames(ps_ww@sam_data),ww.metadata$SampleName)]

ps_ww@sam_data$SiteType <- ww.metadata$SiteType[match(rownames(ps_ww@sam_data),ww.metadata$SampleName)]

ps_ww@sam_data$Urbanization <- ww.metadata$Urbanization[match(rownames(ps_ww@sam_data),ww.metadata$SampleName)]

ps_ww@sam_data$SampleName <-rownames(ps_ww@sam_data)

ps_ww@sam_data$HostSpecies <- ww.metadata$HostSpecies[match(rownames(ps_ww@sam_data),ww.metadata$SampleName)]

ps_ww@sam_data$PISA <- ww.metadata$PISA[match(rownames(ps_ww@sam_data),ww.metadata$SampleName)]

saveRDS(ps_ww, file = "/Tabima_lab/lab_data/Maddie_paper/ps.Rds")
ps_ww <- readRDS(file = "/Tabima_lab/lab_data/Maddie_paper/ps.Rds")

```

### Removing all non-fecal data

```{r}
ps.final <- subset_samples(ps_ww, SampleType == "Fecal") 
saveRDS(ps.final, file = "/Tabima_lab/lab_data/Maddie_paper/ps.final.Rds")
ps.final <- readRDS(file = "/Tabima_lab/lab_data/Maddie_paper/ps.final.Rds")
```


### Total number of samples


```{r}
table(ps.final@sam_data$HostSpecies)
```
***

# Basidiobolus analysis

```{r}
basid <- subset_taxa(ps.final, Phylum=="p__Basidiobolomycota")
dna <- Biostrings::DNAStringSet(taxa_names(basid))
names(dna) <- taxa_names(basid)
basid <- merge_phyloseq(basid, dna)
taxa_names(basid) <- paste0("ASV", seq(ntaxa(basid)))


# Pruning data
basid <- prune_taxa(taxa_sums(basid) > 10, basid) 
basid <- prune_samples(sample_sums(basid) > 30, basid)
basid <- subset_samples(basid, HostSpecies %in% c("Lithobates clamitans", "Lithobates catesbeianus", "Lithobates palustris", "Plethodon cinereus"))

# Saving object
saveRDS(basid, file = "/Tabima_lab/lab_data/Maddie_paper/basid.Rds")
basid <- readRDS(file = "/Tabima_lab/lab_data/Maddie_paper/basid.Rds")
Biostrings::writeXStringSet(basid@refseq, file = "/Tabima_lab/lab_data/Maddie_paper/Basidiobolus_asv.fasta", format = "fasta")
write.table(basid@tax_table, file = "/Tabima_lab/lab_data/Maddie_paper/Taxa_table.txt", quote = F, sep = "\t")
```

## General Basi stats

```{R}
ggplot(data.frame(basid@sam_data) %>% group_by(Urbanization,Site, HostSpecies) %>% tally(), aes(x=Site, y=HostSpecies, label = n, fill=n)) + 
  geom_tile(stat="identity", color = "grey") + 
   geom_text(aes(label = n), color = "white", size = 4) + 
  facet_grid(. ~ Urbanization, scales= "free", space = "free") + 
  theme_classic() +
  scale_fill_continuous(na.value = 'white') + 
  scale_x_discrete(guide = guide_axis(angle = 90))+ ylab("Host Species")
  

data.frame(basid@sam_data) %>% group_by(Site) %>% tally()

```

### Samples with Basi

```{r}
ps.final@sam_data$SampleName[!ps.final@sam_data$SampleName %in% basid@sam_data$SampleName]
```

# Novel ITS

### Extracting unknown ASV's

```{r}
na.basi <- names(basid@tax_table@.Data[,'Species'])[is.na(basid@tax_table@.Data[,'Species'])]
basid.na <- basid@refseq[names(basid@refseq) %in% na.basi]
writeXStringSet(basid.na, filepath = "/Tabima_lab/lab_data/Maddie_paper/Basi_ITS_NA.fasta")
```

#### Adding UNITE Basidiobolus

```{r}
library(seqinr)
unite.db <- read.fasta("/Tabima_lab/raw_data/watershed/fecal/sh_general_release_dynamic_16.10.2022.fasta")
basi.unite <- unite.db[grep(pattern = "Basidiobolus", getName(unite.db))]
basi.unite <- basi.unite[grep(pattern = "_sp", getName(basi.unite),invert = T)]
seqinr::write.fasta(basi.unite, names = getName(basi.unite), file.out = "basi.unite.fasta", nbchar = 10000)
```

```bash
mafft --adjustdirectionaccurately Basi_all.fasta > Basi.alin.fasta
clipkit Basi.alin.fasta -m gappy -g 0.4
```

```{r}
library(msaR)
basi.alin <- read.FASTA("Basi.alin.fasta.clipkit")
msaR(basi.alin)
```

```bash
/Smaug_SSD/bin/iqtree2 -s Basi.alin.fasta.clipkit -B 1000
```

```{r}
library(ggtree)
library(ape)
library(phangorn)

its.tree <- read.tree("Basi.alin.fasta.clipkit.treefile")
its.tree <- root(its.tree, outgroup = "Conidiobolus_sp._ARSEF_7942", resolve.root = T, edgelabel = T)

#its.tree.drop <- drop.tip(its.tree.drop, "Conidiobolus_sp._ARSEF_7942")
p <- ggtree(its.tree) + geom_tiplab(linesize=.5) + geom_nodelab(size=4, vjust=-1, hjust=1)
p
```


### Changing taxonomy in TAXA object

```{r}
ASV_novel <- read.table('ASV_assignment.txt')
table(ASV_novel$V2)
```

### Total per assignment
```{r}
basid@tax_table@.Data[is.na(basid@tax_table@.Data[,'Species']),7] <- ASV_novel$V2
```

## Abundance per OTU 

###per Site

```{r}
## Species
basi.present.df <- as.data.frame(otu_table(basid)@.Data)
basi.present.df[basi.present.df == 0] <- NA
basi.present.df$SampleName <- rownames(basi.present.df)
basi.present.df <- pivot_longer(basi.present.df, !SampleName)
basi.present.df <- basi.present.df[!is.na(basi.present.df$value),]


basi.merged.df <- merge(basi.present.df, sample_data(basid) %>% data.frame, by.x="SampleName")
basi.merged.df <- basi.merged.df[!is.na(basi.merged.df$value),]
basi.merged.df <- tibble(basi.merged.df)

# Site summary
site.summary <- basi.merged.df %>% group_by(Site) %>% dplyr::count(SampleName)
site.summary %>% dplyr::count(Site)

site_diffs <- aov(lm(data = site.summary, formula = n ~ Site))
summary(site_diffs)
TukeyHSD(site_diffs, ordered = T)

box_site <- ggplot(site.summary, aes(x=Site, y=n, fill=Site)) + geom_boxplot() + theme_classic() + scale_fill_grey(start = 0.3) + scale_x_discrete(guide = guide_axis(angle = 90)) + ylab("Number of Basidiobolus ASV's")

# Host summary
site.summary <- basi.merged.df %>% group_by(HostSpecies) %>% dplyr::count(SampleName)

host_diffs<- aov(lm(data = site.summary, formula = n ~ HostSpecies))
summary(host_diffs)
TukeyHSD(host_diffs)
box_host <- ggplot(site.summary, aes(x=HostSpecies, y=n, fill=HostSpecies)) + geom_boxplot() + theme_classic() + scale_fill_grey(start = 0.3) + scale_x_discrete(guide = guide_axis(angle = 90)) + ylab("Number of Basidiobolus ASV's")

# Urbanization summary
site.summary <- basi.merged.df %>% group_by(Urbanization) %>% dplyr::count(SampleName)
urb_diffs <- aov(lm(data = site.summary, formula = n ~ Urbanization))
summary(urb_diffs)
TukeyHSD(urb_diffs)
box_urb <- ggplot(site.summary, aes(x=Urbanization, y=n, fill=Urbanization)) + geom_boxplot() + theme_classic() + scale_fill_grey(start = 0.3) + scale_x_discrete(guide = guide_axis(angle = 90))+ ylab("Number of Basidiobolus ASV's")

# FIGURE 2:
ggarrange(box_site, box_host, box_urb, ncol = 3, nrow = 1, legend = "none")

# Figure 3:
ps.prop <- transform_sample_counts(basid, function(otu) otu/sum(otu))
plot_bar(ps.prop, fill = "Species") +  geom_col(position = "fill") + viridis::scale_fill_viridis(discrete = T) + facet_grid(. ~ Site, scales = "free_x", space = "free_x") + theme(legend.position = 'bottom')

+ theme_bw()
```

Cook	21
Coe's	23
Mill	23
Patch	29
Barre	30
Middle	32
Kinney	44
Moose	44
Tatnuck	50
Thayer	72

## Shannon index

```{r}

richness.basi <- estimate_richness(basid)
anova.sh = aov(richness.basi$Shannon ~ sample_data(basid)$Site)
anova.sh = aov(richness.basi$Shannon ~ sample_data(basid)$HostSpecies)
anova.sh = aov(richness.basi$Shannon ~ sample_data(basid)$Urbanization)
summary(anova.sh)
TukeyHSD(anova.sh)

div.sites <- plot_richness(basid, x = "Site", measures = c("Shannon")) + geom_boxplot() + theme_classic() + scale_fill_grey(start = 0.3) + scale_x_discrete(guide = guide_axis(angle = 90))
div.hosts <- plot_richness(basid, x = "HostSpecies", measures = c("Shannon")) + geom_boxplot() + theme_classic() + scale_fill_grey(start = 0.3)  + scale_x_discrete(guide = guide_axis(angle = 90))
div.urb <- plot_richness(basid, x = "Urbanization", measures = c("Shannon")) + geom_boxplot() + theme_classic() + scale_fill_grey(start = 0.3) + scale_x_discrete(guide = guide_axis(angle = 90))


ggarrange(div.sites, div.hosts, div.urb, cols = 3, nrow = 1, legend = "none")
```

## Beta Diversity


## NMDS
```{r}
ps.prop <- transform_sample_counts(basid, function(otu) otu/sum(otu))
ord.nmds.bray <- ordinate(ps.prop, method="NMDS", distance="bray")

# Sites
site.nmds <- plot_ordination(basid, ord.nmds.bray, color="Site", label="SampleName") + geom_point() + stat_ellipse() + theme_classic() + theme(legend.position = "bottom") + scale_color_viridis(discrete = T, end = 0.8)
adonis2(basid@otu_table ~ basid@sam_data$Site, method='bray', permutations = 999)

# Host
host.nmds <- plot_ordination(basid, ord.nmds.bray, color="HostSpecies", label="SampleName") + geom_point() + stat_ellipse() + theme_classic() + theme(legend.position = "bottom") + scale_color_viridis(discrete = T, end = 0.8)
adonis2(basid@otu_table ~ basid@sam_data$HostSpecies, method='bray', permutations = 999)

# Urbanization
urb.nmds <- plot_ordination(basid, ord.nmds.bray, color="Urbanization", label="SampleName") + geom_point() + stat_ellipse() + theme_classic() + theme(legend.position = "bottom") + scale_color_viridis(discrete = T, end = 0.8)
adonis2(basid@otu_table ~ basid@sam_data$Urbanization, method='bray', permutations = 999)

ggarrange(site.nmds, host.nmds, urb.nmds, cols = 3, nrow = 1)
```

# Geographic separation

## Mantel's test

```{r}
#1. Lat long distance
gs4_deauth()
summer.ww <- read_sheet("https://docs.google.com/spreadsheets/d/1-R5HzaB2c5bHIbekrREiTmvWsC1Y-jhv5wHbcgmy5uc/edit#gid=51094388", sheet = 5)
dist.df <- cbind(summer.ww$Latitude, summer.ww$Longitude)
rownames(dist.df) <- summer.ww$Specimen
dist.df <- dist.df[rownames(dist.df)  %in% rownames(basid@sam_data),]
station.dists <- dist(dist.df)

#2. Community distance
dist.basi.locations <- phyloseq::distance(basid, method="bray", type = "Sites")

# 3 Mantle's test
bray_geo.mantle <- mantel(station.dists, dist.basi.locations, permutations = 10000)
bray_geo.mantle
plot(station.dists, dist.basi.locations)

# 4. Lineal model

gs4_deauth()
metadata <- readxl::read_xlsx("DATA.xlsx")
#metadata <- read_sheet("https://docs.google.com/spreadsheets/d/1-R5HzaB2c5bHIbekrREiTmvWsC1Y-jhv5wHbcgmy5uc/edit#gid=915474950", sheet= 6)

library(lme4)
library(lmerTest)
linear <- lm(Shannon ~ host*Area*Site, data = metadata)
summary(linear)
aov(linear) %>% summary

lineal <- lmer(Shannon ~ Species*Area*Site +  (1|Day), data=metadata)
anova(lineal) 
ranova(lineal)
summary(lineal)
```

Mantel statistic r: 0.06062 
Significance: 0.092291 

## Unique samples DESeq

```{r}

library(DESeq2)
hostdds = phyloseq_to_deseq2(basid, ~ HostSpecies)
gm_mean = function(x, na.rm=TRUE){
  exp(sum(log(x[x > 0]), na.rm=na.rm) / length(x))
}
geoMeans = apply(counts(hostdds), 1, gm_mean)
hostdds = estimateSizeFactors(hostdds, geoMeans = geoMeans)
hostdds = DESeq(hostdds, fitType="local")

res = results(hostdds)
res = res[order(res$padj, na.last=NA), ]
alpha = 0.01
sigtab = res[(res$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(basid)[rownames(sigtab), ], "matrix"))
head(sigtab)

sigtabgen = subset(sigtab, !is.na(Genus))
# Phylum order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Phylum = factor(as.character(sigtabgen$Phylum), levels=names(x))
# Genus order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Genus = factor(as.character(sigtabgen$Genus), levels=names(x))
ggplot(sigtabgen, aes(y=Species, x=log2FoldChange, color=Phylum)) + 
  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
  geom_point(size=6) + 
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
```

## Unique samples counts

### Host species
```{r}
total.basi <- psmelt(basid)

host_asv <- total.basi %>% group_by(HostSpecies, OTU) %>% summarise(sum(Abundance))
host_asv <- pivot_wider(host_asv, names_from = HostSpecies, values_from = `sum(Abundance)`)

# Species and ASV's across all species
asv_all <- cbind(as(host_asv, "data.frame"), as(tax_table(basid)[host_asv$OTU, ], "matrix"))
asv_all <- pivot_longer(asv_all, cols=c('Lithobates catesbeianus','Lithobates clamitans', 'Lithobates palustris', 'Plethodon cinereus'))
asv_all <- asv_all[asv_all$value > 0,]
asv_all <- asv_all %>% group_by(Species, name) %>% dplyr::count(name)
asv_all <- pivot_wider(asv_all, names_from = name, values_from = n)

host_single <- host_asv[apply(host_asv[,-1], 2, function (x) {x == 0}) %>% rowSums == 3,]
host_single <- cbind(as(host_single, "data.frame"), as(tax_table(basid)[host_single$OTU, ], "matrix"))
host_single <- pivot_longer(host_single, cols=c('Lithobates catesbeianus','Lithobates clamitans', 'Lithobates palustris', 'Plethodon cinereus'))
host_single <- host_single[host_single$value > 0,]
host_single <- host_single %>% group_by(Species, name) %>% dplyr::count(name)
pivot_wider(host_single, names_from = name, values_from = n)

ggplot(host_single, aes(x=name, y=n, fill=Species)) + geom_bar(stat="identity", position = "stack")
```

### Urbanization gradient

```{r}
urb.basi <- total.basi %>% group_by(Urbanization, OTU) %>% summarise(sum(Abundance))

urb.basi.sum <- pivot_wider(urb.basi, names_from = Urbanization, values_from = `sum(Abundance)`)
unique.urb <- urb.basi.sum[apply(urb.basi.sum[,-1], 2, function (x) {x == 0}) %>% rowSums == 1,]
```

### Shared betwen urbanization areas
```{r}
shared.urb <- urb.basi.sum[apply(urb.basi.sum[,-1], 2, function (x) {x == 0}) %>% rowSums == 0,]

shared.urb <- cbind(as(shared.urb, "data.frame"), as(tax_table(basid)[shared.urb$OTU, ], "matrix"))
shared.urb$total.ab <- (shared.urb$High + shared.urb$Low)
shared.urb$low.perc <- ((shared.urb$Low/shared.urb$total.ab) * 100) * -1
shared.urb$high.perc <- (shared.urb$High/shared.urb$total.ab) * 100
shared.urb <- pivot_longer(shared.urb, cols=c('low.perc','high.perc'))

 ggplot(shared.urb, aes(x=OTU,fill=name,y=value)) + 
   geom_bar(stat="identity") + 
   geom_hline(aes(yintercept=0)) + 
   facet_grid(. ~ Species, scales = "free_x", space = "free") + 
   theme_classic() + 
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
```

### Unique per urbanization areas
```{r}

urb.summ <- pivot_longer(unique.urb, cols=c('High','Low'))
urb.summ <- urb.summ[!urb.summ$value == 0,]
urb.summ <- cbind(as(urb.summ, "data.frame"), as(tax_table(basid)[urb.summ$OTU, ], "matrix"))

table(urb.summ$name)

ggplot(urb.summ %>% group_by(name, Species) %>% tally(), aes(x=name,y=n,fill=Species)) + geom_bar(stat="identity") + theme_classic() + scale_fill_viridis(discrete = T)
``` 
